Primer on Production Functions and Economic Growth

### 2014-11-28 Articles/Prod Func Primer/ProdFuncPrimer.tex
New ProdFuncPrimer files.

% This article has been prepared for publication in Ecological Economics in RStudio with knitr.
% According to http://www.elsevier.com/author-schemas/the-elsarticle-latex-document-class, we should be using the
% elsarticle.cls file.
% According to http://cdn.elsevier.com/assets/pdf_file/0006/109392/journal_refstyles.pdf, we should be using
% elsarticle-template-2-harv.tex as the template for the text.
% Furthermore, we should be using model2-names.bst for the bibliographic references.
% The approach here is to load the frontmatter and backmatter from elsarticle-template-2-harv.tex
% both ahead of and behind the text for our paper.
% -- Matthew Kuperus Heun, 2014-08-01

@@

\usepackage{wrapfig}    % Allows wrapping of text around figures
\usepackage{soul}       % Provides strikethrough text
\usepackage{float}      % Allows precise positioning of tables and figures within the text
\usepackage{amsmath}    % Allows \begin{equation*} \end{equation*} for unnumbered equations
\usepackage{multirow}   % To create multirow tables
\usepackage{hyperref}   % To create hyperlinks in the paper
\usepackage{microtype}  % produces hanging punctuation and beautiful type

@@

\journal{Ecological Economics}

@@

<<setup, echo=FALSE, message=FALSE, eval=TRUE>>=
require(EconModels)
require(EconData)
require(ggplot2)
require(dplyr) # Load in this order or the mutate and arrange functions will not work as desired.
# Provides access to data melting. 
# See http://cran.r-project.org/web/packages/reshape2/reshape2.pdf
require(reshape2) 
require(knitr)
require(mosaic)

opts_chunk$set(
  # ********************************************************
  eval=TRUE,
  # ********************************************************
  # tikz allows LaTeX code in graphical output. 
  # E.g., "$y$" for a variable name in the legend for a graph.  
  # dev='tikz',
  tidy=FALSE,
  comment=NA,
  # Tells whether to cache output from chunks are saved. Cacheing saves time. 
  # However, references to LaTeX tables DO NOT WORK if cacheing is turned on.
  cache=FALSE,
  warning=FALSE,  #Tells whether to show warnings in the output.
  # Turns off messages for all chunks. 
  # Set TRUE on an individual chunk to see it.
  message=FALSE, 
  # Tells whether to echo code for all chunks. 
  # Set TRUE on an individual chunk to see its code.
  echo=FALSE     
)
@

<<constants, include=FALSE, eval=TRUE>>=
spaghettiGraphLevel <- 1.0 # Spaghetti graph bands will encompass 95% of the data.
@

<<data, include=FALSE>>=
# Load historical data
historicalData <- plyr::rbind.fill(REXS1960, IST, Leeds) # From the EconData package

# Add some pre-econometric data
historicalData <- transform(historicalData, KPerGDP = iK/iGDP)
historicalData <- transform(historicalData, LPerGDP = iL/iGDP)
historicalData <- transform(historicalData, XpPerGDP = iXp/iGDP)
historicalData <- transform(historicalData, XfPerGDP = iXf/iGDP)
historicalData <- transform(historicalData, UPerGDP = iU/iGDP)

# The working directory is set to the directory that contains this file.
# To reach the post-processed data, we need to move 
# up two levels and into the data_postprocessed directory.
# So, in each of the loadPostProcessedData calls below, we use
# dir = "../../data_postprocessed".
pp_path <- file.path("..", "..", "data_postprocessed")

# Load coefficients from all models in one large data frame
REXS1960_Coeffs <- loadPostProcessedData(Source = "REXS1960", kind = "coeffs", dir = pp_path)
IST_Coeffs <- loadPostProcessedData(Source = "IST", kind = "coeffs", dir = pp_path)
Leeds_Coeffs <- loadPostProcessedData(Source = "Leeds", kind = "coeffs", dir = pp_path)
AllCoeffs <- rbind(REXS1960_Coeffs, IST_Coeffs, Leeds_Coeffs)
AllCoeffs$Source <- as.factor(replace(AllCoeffs$Source, which(AllCoeffs$Source=="REXS1960"), "REXS"))
AllCoeffs$Source <- relevelFactor(AllCoeffs$Source, levs = c("REXS", "IST", "Leeds"))
AllCoeffs$energy <- relevelFactor(AllCoeffs$energy, levs = energyLevels)
# Load fitted data from all models in one large data frame
REXS1960_Fitted <- loadPostProcessedData(Source = "REXS1960", kind = "fitted", dir = pp_path)
IST_Fitted <- loadPostProcessedData(Source = "IST", kind = "fitted", dir = pp_path)
Leeds_Fitted <- loadPostProcessedData(Source = "Leeds", kind = "fitted", dir = pp_path)
AllFitted <- rbind(REXS1960_Fitted, IST_Fitted, Leeds_Fitted)
AllFitted$Source <- replace(AllFitted$Source, which(AllFitted$Source=="REXS1960"), "REXS")
AllFitted$Source <- relevelFactor(as.factor(AllFitted$Source), levs = c("REXS", "IST", "Leeds"))
AllFitted$energy <- relevelFactor(AllFitted$energy, levs=energyLevels)

# Load models
REXS1960_Models <- loadPostProcessedData(Source = "REXS1960", kind = "models", dir=pp_path)
IST_Models <- loadPostProcessedData(Source = "IST", kind = "models", dir=pp_path)
Leeds_Models <- loadPostProcessedData(Source = "Leeds", kind = "models", dir=pp_path)
AllModels <- unlist(list(REXS1960_Models, IST_Models, Leeds_Models), recursive=FALSE)
# The function to extract data from the models
extract_data <- function(x, name) {
  # Names are in the form Source.Country.modelType.factor_string
  # Parse various pieces of name
  name_pieces <- strsplit(name, split=".", fixed=TRUE)[[1]]
  Source <- name_pieces[[1]]
  Country <- name_pieces[[2]]
  modelType <- name_pieces[[3]]
  factorString <- name_pieces[[4]]
  # Parse the factorString to get the energyType
  energyType <- parseFactorString(factorString)$energyType
  data.frame(year = (attr(x, which="data"))$Year, resid = resid(x), name=name, Source=Source, Country=Country,
             modelType=modelType, energyType=energyType)
}
residual_data <- leaf_apply(AllModels, extract_data, class=c("CDEmodel", "cesEst", "LINEXmodel"))
# Bind these all in a data frame. 
residuals_df <- do.call(rbind, residual_data)
# Re-order the factors as desired for our graphs.
residuals_df$Source <- as.factor(
  replace(as.character(residuals_df$Source), which(residuals_df$Source=="REXS1960"), "REXS"))
residuals_df$Source <- relevelFactor(residuals_df$Source, levs = c("REXS", "IST", "Leeds"))
residuals_df$energyType <- as.factor(
  replace(as.character(residuals_df$energyType), which(is.na(residuals_df$energyType)), "none"))
residuals_df$energyType <- relevelFactor(as.factor(residuals_df$energyType), levs = energyLevels)
@

% These equations are used in several locations in the paper.
% By defining commands for them, we ensure consistent appearance
% of the equations when they are used.

% Add an argument if you want a label. Example: "\label{eq:CES}"
\newcommand{\ceskl}[1]{\begin{equation} #1
  y = \gamma \: A \: \left[\delta_1 k^{-\rho_1} 
      + (1-\delta_1)l^{-\rho_1} \right]^{-1/\rho_1}; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}}

\newcommand{\ceskle}[1]{\begin{equation} #1
  y = \gamma \: A \: \left\{\delta \left[\delta_1 k^{-\rho_1} 
      + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
      + (1-\delta) e^{-\rho} \right\}^{-1/\rho}; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}
}

\newcommand{\ceslek}[1]{\begin{equation} #1
  y = \gamma \: A \: \left\{\delta \left[\delta_1 l^{-\rho_1} 
      + (1-\delta_1) e^{-\rho_1} \right]^{\rho/\rho_1} 
      + (1-\delta) k^{-\rho} \right\}^{-1/\rho}; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}
}

\newcommand{\cesekl}[1]{\begin{equation} #1
  y = \gamma \: A \: \left\{\delta \left[\delta_1 e^{-\rho_1} 
      + (1-\delta_1) k^{-\rho_1} \right]^{\rho/\rho_1} 
      + (1-\delta) l^{-\rho} \right\}^{-1/\rho}; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}
}

@@

\title{Paper 2 !DRAFT! Do not circulate!}

@@

% \author[BenWarr]{Benjamin Warr}
% \author[IST]{Tiago Domingos}
% \author[Leeds]{Paul Brockway}
% \author[Leeds]{Julia Steinberger}
% \author[CalvinEngr]{Matthew Kuperus Heun\corref{cor1}}
% \ead{mkh2@calvin.edu}

% \address[BenWarr]{Ben Warr's address, Zambia}
% \address[IST]{IST, Lisbon, Portugal}
% \address[Leeds]{Leeds University, Leeds, UK}
% \cortext[cor1]{Corresponding author}
% \address[CalvinEngr]{Engineering Department, Calvin College, Grand Rapids, MI 49546, USA}

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
economic growth \sep exergy \sep energy \sep Cobb-Douglas \sep CES \sep LINEX
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Coordinates of Analysis} 
\label{sec:Coordinates}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section describes the coordinates of analysis and briefly reviews 
literature related to each.

%++++++++++++++++++++++++++++++
\subsection{Mathematical Forms of the Energy-augmented Production Function} 
\label{sec:Prod_Func_Forms}
%++++++++++++++++++++++++++++++

In this paper, we assess three prominent energy-augmented production functions 
that appear in the literature: 
Cobb-Douglas (CD), 
Constant Elasticity of Substitution (CES), and 
LINear EXponential (LINEX). 
The following subsections describe each.

%------------------------------
\subsubsection{Cobb-Douglas Production Function} 
\label{sec:CDe}
%------------------------------

The Cobb-Douglas production function can be expressed as
%
\begin{equation} \label{eq:CD}
  y = \theta \mathrm{e}^{\lambda(t-t_0)} k^\alpha l^\beta \;,
\end{equation}
%
\noindent where 
$y \equiv Y/Y_{0}$,
$\theta$ is a scale parameter,
e is the base of the natural logarithm, 
$\lambda$ is represents the pace of technological progress,
$t$ (time) is measured in years,
$k \equiv K/K_{0}$, 
$l \equiv L/L_{0}$, 
$Y$ (economic output) is represented by GDP, 
$K$ (capital) is expressed in currency units, 
$L$ (labor) is expressed in workers or work-hours/year, and
the 0 subscript indicates values at an initial year.%
  \footnote{Dimensionless, indexed quantities are represented by 
  lower-case symbols 
  ($y$, $k$, $l$, $e$, $q$, $x$, and $u$), and dimensional 
  quantities are represented by upper-case symbols 
  ($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$). 
  Model parameters are represented by Greek letters
  ($\alpha$, $\beta$, $\lambda$, $\theta$).
  }
Constant returns to scale are represented by the constraint
$\alpha + \beta = 1$.

The capital-labor Cobb-Douglas production function shown in Equation~\ref{eq:CD}
can be augmented to include an energy term:
%
\begin{equation} \label{eq:CDe}
  y = \theta \mathrm{e}^{\lambda(t-t_0)}  k^\alpha l^\beta e^\gamma \;,
\end{equation}
%
where $e \equiv E/E_0$, and $E$ is in units of energy per time, typically TJ/year. 
The energy-augmented Cobb-Douglas production function 
is often assumed to have constant returns to scale for the three factors 
of production: $\alpha + \beta + \gamma = 1$.

The term $\mathrm{e}^{\lambda (t-t_0)}$ is known as total factor productivity~($A$),
and $\lambda$ is the Solow residual. 
Assuming that parameters $\theta$, $\alpha$, and $\beta$ are known from parameter estimation,
we can estimate the value of $\lambda$ as follows.
First, assume constant returns to scale such that $\gamma = 1 - \alpha - \beta$.
Then, take the natural logarithm ($\ln$) of Equation~\ref{eq:CD} to obtain
%
\begin{equation} \label{eq:lnCD}
  \ln y = \ln \theta + \ln A + \alpha \ln k + \beta \ln l + (1 - \alpha - \beta) \ln e \;.
\end{equation}
%
By taking the derivative of Equation~\ref{eq:lnCD} with respect to time ($t$) 
and noting that model parameters $\theta$, $\alpha$, and $\beta$ are 
constant with respect to time, we obtain
%
\begin{equation} \label{eq:ddtlnCD}
  \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      = \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      + \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} \;.
\end{equation}
%
Solving for the Solow Residual term gives
%
\begin{equation} \label{eq:solow_resid}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
        \right] \;.
\end{equation}
%
Recognizing that $A \equiv \mathrm{e}^{\lambda (t-t_0)}$ gives
%
\begin{equation}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
    = \frac{1}{\mathrm{e}^{\lambda (t-t_0)}} \; \lambda \mathrm{e}^{\lambda (t-t_0)}
    = \lambda \; ,
\end{equation}
%
which can be substituted into Equation~\ref{eq:solow_resid} to find
%
\begin{equation} \label{eq:lambda}
  \lambda 
    = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
    - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
      \right] \;.
\end{equation}
%
Equation~\ref{eq:lambda} applies for any instant in time.

If we approximate derivatives in Equation~\ref{eq:lambda} with forward differences
between times $i$ and $j$, we find
%
\begin{equation} \label{eq:lambdaij}
  \lambda_{i,j}
    \approx \frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i} 
    - \left[
            \alpha \frac{1}{k_i}\frac{k_j - k_i}{t_j - t_i} 
            + \beta \frac{1}{l_i}\frac{l_j - l_i}{t_j - t_i} 
            + (1 - \alpha - \beta) \frac{1}{e_i}\frac{e_j - e_i}{t_j - t_i} 
      \right] \;,
\end{equation}
% 
where $\lambda_{i,j}$ approximates the true, instantaneous value of $\lambda$
between times $i$ and $j$.

Interestingly, 
%
\begin{equation}
  \lambda_{1,n}
    \neq \sum_{i=2}^{n} \lambda_{i,i-1}  \;.
\end{equation}
% 
Rather,
%
\begin{equation} \label{eq:lambda1n}
  \lambda_{1,n}
    \approx \frac{1}{y_1}\frac{y_n - y_1}{t_n - t_1} 
      - \left[ 
            \alpha \frac{1}{k_1}\frac{k_n - k_1}{t_n - t_1} 
            + \beta \frac{1}{l_1}\frac{l_n - l_1}{t_n - t_1} 
            + (1 - \alpha - \beta) \frac{1}{e_1}\frac{e_n - e_1}{t_n - t_1} 
      \right] \; ,
\end{equation}
% 
which will become increasingly inaccurate over large time spans,
because $y_1$, $k_1$, $l_1$, and $e_1$ will be less representative
of the average value of $y$, $k$, $l$, and $e$ in the time span, respectively.
This suggests that an averaging approach such as
%
\begin{equation}
  \lambda_{1,n}
    \approx \frac{\sum_{i=2}^{n} \lambda_{i,i-1}}{n-1}
\end{equation}
%
is a better approximation of Equation~\ref{eq:lambda1n},
which is, itself, an approximation of the true value of $\lambda$ 
given by Equation~\ref{eq:lambda}.

%------------------------------
\subsubsection{Constant Elasticity of Substitution Production Function (CES)} 
\label{sec:CES}
%------------------------------

Other energy economists use an energy-augmented 
Constant Elasticity of Substitution (CES) production function to 
describe economic growth. The \texttt{R} \citep{R} package 
\texttt{micEconCES} \citep{Henningsen:2011td} estimates CES production 
functions of the following forms (among others):
%
\ceskl{\label{eq:CES}}
%
\ceskle{\label{eq:CESkle}}
%
\noindent Equation~\ref{eq:CES} is a CES production function with 
capital stock ($k$) and labor ($l$) factors of production. 
Equation~\ref{eq:CESkle} augments Equation~\ref{eq:CES} with energy 
using a ($kl$)($e$) nesting structure, as is typical in the literature. 
Equation~\ref{eq:CES} is a degenerate form of Equation~\ref{eq:CESkle} 
where $\delta \rightarrow 1$.

$\gamma$ is a fitting parameter that represents productivity. 
The fitting parameters $\rho_1$ and $\rho$ indicate elasticities 
of substitution ($\sigma_1$ and $\sigma$).
The elasticity of substitution 
between capital ($k$) and labor ($l$) is given by 
$\sigma_1 = \frac{1}{1+\rho_1}$, and
the elasticity of substitution between ($kl$) and ($e$) is given by 
$\sigma = \frac{1}{1+\rho}$. 
As $\rho_1 \rightarrow 0$, $\sigma_1 \rightarrow 1$,
and the embedded CES production function for $k$ and $l$ degenerates 
to the Cobb-Douglas production function.
Similarly, as $\rho \rightarrow 0$, $\sigma \rightarrow 1$,
and the CES production function for $(kl)$ and $(e)$ degenerates 
to the Cobb-Douglas production function.
As $\sigma \rightarrow \infty$ ($\rho \rightarrow -1$), 
($kl$) and ($e$) are perfect substitutes. 
As $\sigma \rightarrow 0$ ($\rho \rightarrow \infty$), 
($kl$) and ($e$) are perfect complements: 
no substitution is possible. Similarly, 
as $\sigma_1 \rightarrow 0$ ($\rho_1 \rightarrow \infty$),
$k$ and $l$ are perfect complements.
$\delta_1$ describes the relative importance of capital ($k$)
and labor ($l$), and
$\delta$ describes the importance of ($kl$) relative to ($e$).

Constraints on the fitting parameters include 
$\delta_1 \in [0,1]$,
$\delta \in [0,1]$,
$\rho_1 \in [-1,0) \cup (0,\infty)$, and
$\rho \in [-1,0) \cup (0,\infty)$.

Two other nestings of the factors of production ($k$, $l$, and $e$)
are possible with the CES model.
%
\ceslek{\label{eq:CESlek}}
%
\cesekl{\label{eq:CESekl}}
%
\noindent Note that the $\rho$ ($\sigma$) and $\delta$ parameters have different
meanings that depend upon the nesting of the factors of production.

%------------------------------
\subsubsection{LINEX Production Function} 
\label{sec:LINEX}
%------------------------------

A third production function, the LINear EXponential (LINEX) function
%
\begin{equation} \label{eq:LINEX}
  y = \theta \; \mathrm{e}^{a_0\left[2 \left(1 - \frac{1}{\rho_k} \right) 
    + c_t \left(\vphantom{\frac{1}{\rho_k}}\rho_l - 1 \right)\right]} \; e \;,
\end{equation}
%
was derived from thermodynamic considerations
\citep{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. 
In contrast to both the Cobb-Douglas model (Equations~\ref{eq:CD} and 
\ref{eq:CDe}) and the CES model (Equations~\ref{eq:CES}--\ref{eq:CESekl}), 
the LINEX model (Equation~\ref{eq:LINEX}) does not include a generic, 
time-dependent augmentation term ($A$).
The LINEX model assumes energy ($e$) is the only factor of production.

>>>>

@@

\usepackage{wrapfig}    % Allows wrapping of text around figures
\usepackage{float}      % Allows precise positioning of tables and figures within the text
\usepackage{amsmath}    % Allows \begin{equation*} \end{equation*} for unnumbered equations
\usepackage{multirow}   % To create multirow tables
\usepackage{hyperref}   % To create hyperlinks in the paper
\usepackage{microtype}  % Produces hanging punctuation and beautiful type
\usepackage{booktabs}   % For nice tables

@@

\journal{}

@@

@@

\title{Primer on Production Functions and Economic Growth}

@@

% \author[BenWarr]{Benjamin Warr}
% \author[IST]{Tiago Domingos}
% \author[Leeds]{Paul Brockway}
% \author[Leeds]{Julia Steinberger}
\author[CalvinEngr]{Matthew Kuperus Heun\corref{cor1}}
\ead{mkh2@calvin.edu}

% \address[BenWarr]{Ben Warr's address, Zambia}
% \address[IST]{IST, Lisbon, Portugal}
% \address[Leeds]{Leeds University, Leeds, UK}
\cortext[cor1]{Corresponding author}
\address[CalvinEngr]{Engineering Department, Calvin College, Grand Rapids, MI 49546, USA}

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
production function \sep economic growth \sep energy \sep Cobb-Douglas \sep CES \sep LinEx
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
\end{keyword}

@@

This primer on production functions 
provides background on several production functions
for the purpose of accounting for the role of energy
in economic growth.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Differential Economic Growth} 
\label{sec:diff_growth}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The differential equation for economic growth~($y$) with 
capital~($k$), labor~($l$), and energy~($e$) factors of production, 
i.e., $y = y(k, l, e; t)$ is

\begin{equation} \label{eq:diff_growth}
  \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      = \lambda \frac{1}{t - t_0}
      + \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
      + \gamma \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} \; ,
\end{equation}
%
where 
$y \equiv Y/Y_{0}$,
$t$ (time) is often measured in years,
$\lambda$ represents the time dependence of the change of $y$,
$k \equiv K/K_0$, 
$l \equiv L/L_0$, 
$e \equiv E/E_0$.
$Y$ (economic output) is measured by GDP in currency units, 
$K$ (capital) is expressed in currency units, 
$L$ (labor) is expressed in workers or work-hours/year, and
the subscript 0 indicates values at an initial year.
Constant returns to scale are represented by the constraint
$\alpha + \beta + \gamma = 1$.

The output elasticities $\alpha$, $\beta$, and $\gamma$ are defined as

\begin{equation} \label{eq:alaph_def}
  \alpha \equiv \frac{k}{y} \frac{\partial k / \partial t}{\partial y / \partial t} \; , \;
  \beta  \equiv \frac{l}{y} \frac{\partial k / \partial t}{\partial y / \partial t} \; , \;
  \gamma \equiv \frac{e}{y} \frac{\partial e / \partial t}{\partial y / \partial t} \; .
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Leontief Production Function} 
\label{sec:Leontief}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Leontief production function is given by 
%
\begin{equation} \label{eq:Leontief}
  y = \min \left[ a k, b l, c e\right]
\end{equation}
%
where $a$, $b$, and $c$ are technologically-determined constants. 
The Leontief function is a recipe for production.

An example of a Leontief system is the production of automobiles.
If $Y$ is the production rate of cars (cars/year),
$W$ is wheels (wheels/year), 
$S$ is steering wheels (steering wheels/year), and
$R$ is rear-view mirrors (rear-view mirrors/year), 
a Leontief production function is
%
\begin{equation} \label{eq:cars}
  Y = \min \left[ a W, \; b S, \; c R\right] \; ,
\end{equation}
%
where $a$ = 1 car/4 wheels, $b$ = 1 car/steering wheel, and $c$ = 1 car/rear-view mirror.

The Leontief production function indicates that 
oversupplying a factor of production~($W$, $S$, or $R$) will not increase output~($Y$).
Put another way, wheels~($W$), steering wheels~($S$), and rear-view mirrors~($R$)
are not substitutable.

In the Leontief world, there is always at least one limiting factor of production.
With 8 wheels, 10 steering wheels, and 10 rear-view mirrors,
an automaker can produce only 2 cars.
Under these conditions, wheels are the limiting factor of production, 
and there will be 8 unused steering wheels and 8 unused rear-view mirrors, 
a non-optimal production system indeed.
The optimal (profit-maximizing) production strategy to make 2 cars would be to supply
8 wheels, 2 steering wheels, and 2 rear-view mirrors.

In modern economies, 
constraints exist in terms of capital~($k$), labor~($l$), and energy~($e$).~\citep{Kummel:2010vz}
%
\begin{itemize}

  \item Given a fully-utilized machine, throwing more energy at it will \emph{not} increase output~($y$).%
  \footnote{
  In fact, adding additional energy to the machine may cause damage.
  }
  Capital~($k$) constrains energy~($e$).

  \item Given a state of automation of the economy at an instant in time, 
  and assuming full utilization of the capital~($k$)
  in terms of both labor~($l$) and energy~($e$) for operation,
  adding labor to the economy doesn't increase output~($y$). 
  Capital~($k$) and energy~($e$) constrain labor~($l$).

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear Production Function} 
\label{sec:linear}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The linear production function is given by
%
\begin{equation} \label{eq:linear}
  y = a k + b l + c e \; .
\end{equation}
%
In Equation~\ref{eq:linear}, $a$, $b$, and $c$ are positive constants.
In the linear world, factors of production are perfect substitutes.

Continuing the cooking analogy, consider producing toast ($Y$) 
with a fatty substance~(butter or margarine).
%
\begin{equation} \label{eq:toast}
  Y = a B + b M
\end{equation}
%
The linear model (Equation~\ref{eq:toast}) indicates that butter~($B$) and margarine~($M$) 
can be used interchangably as the fatty substance.
Butter~($B$) and margarine~($M$) are perfect substitutes.%
  \footnote{
  Butter and margarine may be perfect substitutes for all but the most discerning toast consumers.
  }%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cobb-Douglas Production Function} 
\label{sec:CD}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Cobb-Douglas production function is given by
%
\begin{equation} \label{eq:CD}
  y = A k^\alpha l^\beta e^\gamma \; ; \; A \equiv \mathrm{e}^{\lambda(t-t_0)} \; .
\end{equation}
%
$A$ is known as total factor productivity,
and $\lambda$ is the Solow residual.

In the Cobb-Douglas world, factors of production are substitutable, 
eg., a reduction of labor~($l$) can be compensated by an increase of capital~($k$)
without affecting output~($y$).%
  \footnote{
  In fact, the capital-for-labor substitution has a name: automation.
  }
Furthermore, an increase in one factor of production (say, $k$)
without a concomitant increase in the others ($l$ or $e$) 
will result in an increase of output ($y$), provided that the output elasticity 
($\alpha$) is non-zero.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constant Elasticity of Substitution Production Function (CES)} 
\label{sec:CES}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Constant Elasticity of Substitution production function
provides smooth transition between perfectly complementary (Leontief) 
and perfectly substitutable (linear) factors of production.

%++++++++++++++++++++++++++++++
\subsection{CES with Two Factors of Production} 
\label{sec:CES-2}
%++++++++++++++++++++++++++++++

With two factors of production (capital, $k$ and labor, $l$), 
the CES production function is given by
%
\begin{equation} \label{eq:CES}
  y = A \: \left[\delta_1 k^{-\rho_1} 
      + (1-\delta_1)l^{-\rho_1} \right]^{-1/\rho_1} \; ; \;
      A = \mathrm{e}^{\lambda (t-t_0)}   \; .
\end{equation}
%
In Equation~\ref{eq:CES}, capital~($k$) and labor~($l$) are said to be ``nested''
together in the production function.

$\delta_1$ describes the importance of capital ($k$)
relative to labor ($l$).
As $\delta_1 \rightarrow 1$, only capital~($k$) is necessary for production.
As $\delta_1 \rightarrow 0$, only labor~($l$) is necessary for production.

$\rho_1$ indicates the elasticity 
of substitution~($\sigma_1$) between capital~($k$) and labor~($l$).
The elasticity of substitution 
between capital~($k$) and labor~($l$) is related to $\rho_1$ and given by 
%
\begin{equation} \label{eq:sigma_1}
 \sigma_1 = \frac{1}{1+\rho_1} \; .
\end{equation}
%
Solving for $\rho_1$ gives
%
\begin{equation} \label{eq:rho_1}
  \rho_1 = \frac{1 - \sigma_1}{\sigma_1} \; .
\end{equation}

Constraints on the parameters include 
$\delta_1 \in [0,1]$ and
$\rho_1 \in (-1,0) \cup (0,\infty)$.

%++++++++++++++++++++++++++++++
\subsection{CES with Three Factors of Production} 
\label{sec:CES-3}
%++++++++++++++++++++++++++++++

A doubly-nested form of the CES production function is required
if a third factor of production~($e$) is to be included.
%
\begin{equation} \label{eq:CES-kle}
  y = A \: \left\{\delta \left[\delta_1 k^{-\rho_1} 
      + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
      + (1-\delta) e^{-\rho} \right\}^{-1/\rho} \; ; \; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}
%
In Equation~\ref{eq:CES-kle}, $\delta$ 
describes the importance of the capital-labor ($kl$) nesting 
relative to energy~($e$).

Important questions arise: why nest capital~($k$) and labor~($l$) together 
and separate energy~($e$)
in the~($kl$)($e$) nest structure?
Why not nest, e.g., labor~($l$) and energy~($e$) in a~($le$)($k$) structure?%
  \footnote{
  Note that the $\rho$ (and, therefore, $\sigma$) and $\delta$ parameters have different
  meanings depending upon how the factors of production are nested.
  }%
%
\begin{equation} \label{eq:CES-lek}
  y = A \: \left\{\delta \left[\delta_1 l^{-\rho_1} 
      + (1-\delta_1) e^{-\rho_1} \right]^{\rho/\rho_1} 
      + (1-\delta) k^{-\rho} \right\}^{-1/\rho} \; ; \; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}
%
Or nest energy and capital in a~($ek$)($l$) structure?
%
\begin{equation} \label{eq:CES-ekl}
  y = A \: \left\{\delta \left[\delta_1 e^{-\rho_1} 
      + (1-\delta_1) k^{-\rho_1} \right]^{\rho/\rho_1} 
      + (1-\delta) l^{-\rho} \right\}^{-1/\rho} \; ; \; 
      A = \mathrm{e}^{\lambda (t-t_0)}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Relating CES to Other Production Functions} 
\label{sec:CES-others}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Equation~\ref{eq:CES} simplifies to the other production functions 
with different values of $\sigma$.
For the two-factor CES production function, 
Table~\ref{tab:ces_simplifications} describes the options.

\begin{table}
\caption[CES Simplifications]{CES Simplifications.}
\begin{center}
 \begin{tabular}{ c c c c }
\toprule 
$\sigma$ & $\rho$ & CES simplifies to \dots & Factors of production are \dots \\
\midrule
0        & $\infty$ & Leontief              & perfect complements             \\
% \midrule
1        & 0        & Cobb-Douglas          & substitutable                   \\
% \midrule
$\infty$ & $-1$     & linear                & perfect substitutes             \\
\bottomrule
\end{tabular}
\end{center}
\label{tab:ces_simplifications}
\end{table}

Figure~\ref{fig:CES_isoquants} shows the simplifications from Table~\ref{tab:ces_simplifications}
graphically.
%
\begin{figure}[!ht]
\centering\
\includegraphics[width=0.8\linewidth]{figure/CESIsoquants.png}
\caption[CES isoquants]{CES isoquants (lines of constant output). 
$K$ is capital and $N$ is labor.~\citep{Klump:2011aa}}
\label{fig:CES_isoquants}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LinEx Production Function} 
\label{sec:linex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

An additional production function, the Linear Exponential (LinEx%
    \footnote{
    The name ``LinEx'' comes from the form of the equation:
    production is \emph{linear} in energy~($e$) and \emph{exponential}
    in ratios of the factors of production~($k$, $l$, and $e$).
    }%
) function is given by K\"{u}mmel in several places as
%
\begin{equation} \label{eq:LinEx-kummel}
  y = e \exp{\left[ a \left( 2 - \frac{l + e}{k} \right)  + a c \left( \frac{l}{e} - 1 \right)\right]}
\end{equation}
%
An alternative presentation that affords easier interpretation is
%
\begin{equation} \label{eq:LinEx}
  y = A e \; ; \; A \equiv \mathrm{e}^{a\left[2 \left(1 - \frac{1}{\rho_k} \right) 
    + c \left(\vphantom{\frac{1}{\rho_k}}\rho_l - 1 \right)\right]} \; .
\end{equation}
%
The LinEx function was derived from thermodynamic considerations
\citep{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. 
A key feature of the derivation involves
constraints among capital~($k$), labor~($l$), and energy~($e$).
These constraints were discussed in Section~\ref{sec:Leontief} above.%
  \footnote{
  As an aside, I'm not sure the constraints 
  described in Section~\ref{sec:Leontief} are the right ones.
  }%
